---
title: "Compare CFU and FungaLight for log dilution of H2O2"
author: Hanxi Tang, Bin He
date: "2024-03-22 (updated `r Sys.time()`)"
output:
  html_notebook:
    toc: true
    toc_depth: 4
    code_folding: hide
---

# Background
This is the flow cytometry data for post treatment phase C. glabrata on 2024.02.14-17, stained with PI/PI+SYTO9, and run through flow cytometry (details in ELN), and plated for CFU. The goal is to establish FungaLight's distinguishing power and correlate with plating based method. A similar experiment was done for S. cerevisiae on 2024.02.25-28

```{r setup, message=FALSE}
require(tidyverse)
require(flowCore)
require(flowClust)
require(openCyto)
require(ggcyto)
require(cowplot)
require(ggrdiges)
```

```{r}
old <- theme_set(theme_minimal())
```

# Import data
> FCS files are stored in rdss/Users/Hanxi/flow cytometry under the above dates, FCS is read and write into an input tsv table. The tsv file is avaliable from the Input folder.

```{r}
# use relative path to make it easier for collaboration
data.path = "../input/20240214-17 Log H2O2 correlation Cg/"
dat0 <- read.flowSet(path = data.path, pattern = "*.fcs",
                     transformation = FALSE,  # the original values are already linearized. 
                     emptyValue = FALSE,  alter.names = TRUE,   # change parameter names to R format
                     column.pattern = ".H|FSC|SSC") # only load the height variables for the fluorescent parameters
```

Simplify the sample names

```{r}
oriNames <- sampleNames(dat0)
tmp <- str_split(oriNames, pattern = "[ _]+", simplify = TRUE)[,c(1, 6, 7)]
colnames(tmp) <- c("Date", "Treatment", "Dye")
treatment <- c("mock", "5C", "10", "100", "1000")
sample <- data.frame(tmp) %>% 
  mutate(
    Dye = ifelse(Dye == "p.fcs", "PI", ifelse(Dye == "b.fcs", "Both", Dye)), 
    Treatment = factor(Treatment, levels = c(0, "5C", 10, 100, 1000),
                       labels = treatment),
    name = paste(Date, paste0(Treatment, 
                              ifelse(Treatment %in% c("mock", "5C"), "", " mM")), 
                 Dye, sep = "_")
  )
rownames(sample) <- oriNames
dat <- dat0 # make a copy
pData(dat) <- sample
print(pData(dat))
write_tsv(pData(dat), file = "../input/20240214-log-h2o2-cg-sample.tsv")
```


# Gatting for Singlets
>The following gaphing steps are used to gate singlets by FSC and SSC values. Only singlets are included in analysis.
>Gate for outlier 

```{r}
outlier.gate <- rectangleGate(filterId = "-outlier", "FSC.H" = c(1.2e5, 1e6), "SSC.H" = c(1e2, 1e6))
ggcyto(dat[1], aes(x = FSC.H, y = SSC.H), subset = "root") +
  geom_hex(bins = 64) + geom_gate(outlier.gate) + facet_wrap(~name, ncol = 2) + ggcyto_par_set(limits = "instrument")
```
Add gate to GS
```{r}
gs <- GatingSet(dat) # create a GatingSet
# rename the samples
sampleNames(gs) <- pData(gs)$name
# repair the Date column in pData
pData(gs)$Date = paste0("0", pData(gs)$Date)
# add the outlier gate
gs_pop_add(gs, outlier.gate, parent = "root")
recompute(gs)
```

>Gate for singlets

```{r fig.width=4, fig.height=4}
scPars <- ggcyto_par_set(limits = list(x = c(0,1e6), y = c(30,300)))
ex <- Subset(dat[[1]], outlier.gate)
polygon <- matrix(c(1e5, 1e5, 1e6, 1e6, 
                    60, 105, 135,60), ncol = 2)
colnames(polygon) <- c("FSC.H", "FSC.W")
singlet.gate <- polygonGate(filterId = "singlet", boundaries = polygon)
ggcyto(ex, aes(x = FSC.H, y = FSC.W)) + geom_hex(bins = 128) + geom_gate(singlet.gate) + geom_stats() + scPars
```


Add this gate to the gatingSet
```{r}
gs_pop_add(gs, singlet.gate, parent = "-outlier", name = "singlet")
recompute(gs)
```

# Gate for scoring {.tabset}

## Live = red negative

This is the live population with red fluorescence below 10^2.2

> <span style="color:red">Important: I changed the Red top boundary to 10^2.3 from 10^2.2</span>

> **Main takeaway**
>
> 1. From 10mM to 1M treated samples, %live decreases and the change is much more obvious in the dual stain (PI first or simultaneous) than PI alone
>
> 2. The mock treated sample has a visible high red high green (not as high and prominent as the 100 mM treated) than 10 mM treated. It doesn't follow the trend above.

```{r, fig.width=6, fig.height=4}
#scPars <- ggcyto_par_set(limits = list(x = c(0,1e6), y = c(0,1e5)))
polygon <- matrix(c(0, 5*10^3, 5*10^3, 0,
                    10^2.2, 10^2.2, 0, 0), ncol = 2)
colnames(polygon) <- c("BL1.H", "BL3.H")
live.gate <- polygonGate(filterId = "live", .gate = polygon)
for(date in c("021424", "021624", "021724")){
  p <- ggcyto(gs[pData(gs)$Date == date], aes(x = BL1.H, y = BL3.H), subset = "singlet") + 
    geom_hex(bins = 128) + 
    geom_gate(live.gate) + 
    geom_stats(location = "data", adjust = c(0.1, 0.8), digits = 2) + 
    labs(title = paste0("Date: ", date)) +
    facet_grid(Dye ~ Treatment) +
    scale_x_logicle() + scale_y_logicle() + 
    theme(strip.text = element_text(size = rel(1.1), color = "steelblue", face = 2))
  print(p)
}
```

Add this gate to the gatingSet

```{r eval=FALSE, include=FALSE}
gs_pop_add(gs, live.gate, parent = "singlet", name = "live")
recompute(gs)
```
## Intermediate = high G/R
This is the intermediate population with high Red and Green. We hypothesize that they represent the 

> <span style="color:red">Important: I changed the bottom boundary from 10^2.2 to 10^2.3, and the left boundary from 10^3 to 2*10^3</span>

> **Main takeaway**
>
> 1. Different from the %live gate, the intermediate population is most obvious in the 
> 100 mM treated sample, and absent in both the 10 mM and 1M treated ones. This suggests
> that this population could be useful for distinguishing stress levels below the lethal concentration
> but above the tolerance levels of the species - in this case, < 1M and > 10 mM.
>
> 2. The mock treated sample showed an intermediate population. Was unexpected.

```{r, fig.width=6, fig.height=3.5}
#scPars <- ggcyto_par_set(limits = list(x = c(0,1e6), y = c(0,1e6)))
#ex <- Subset(dat.f1g[[9]], singlet.gate)
polygon <- matrix(c(10^5, 1*10^3, 10^5,
                    10^2.2, 10^2.2, 10^4), ncol = 2)
colnames(polygon) <- c("BL1.H", "BL3.H")
inter.gate <- polygonGate(filterId = "inter", .gate = polygon)
for(date in c("021424", "021624", "021724")){
  p <- ggcyto(gs[pData(gs)$Date == date], aes(x = BL1.H, y = BL3.H),
              subset = "singlet") + 
    geom_hex(bins = 128) + 
    geom_gate(inter.gate) + 
    geom_stats(location = "data",  adjust = c(0.1, 0.8), digits = 2) + 
    #geom_gate(live.gate) + 
    labs(title = paste0("Date: ", date)) +
    facet_grid(Dye ~ Treatment) +
    scale_x_logicle() + scale_y_logicle() + 
    theme(strip.text = element_text(size = rel(1.1), color = "steelblue", face = 2))
  print(p)
}
```

```{r eval=FALSE, include=FALSE}
gs_pop_add(gs, inter.gate, parent = "singlet", name = "intermediate")
recompute(gs)
```
# Exported gate stats

% of events in each of the three gates
```{r}
gated_stats <- gs_pop_get_stats(gs, type = "count") %>% 
  as_tibble() %>% 
  mutate(pop = gsub(".*/", "", pop), pop = gsub("-outlier", "cells", pop)) %>% 
  pivot_wider(names_from = pop, names_prefix = "n_", values_from = count) %>% 
  mutate(
    #p_singlet = n_singlet / n_cells,
    p_live = n_live / n_singlet,
    p_intermediate = n_intermediate / n_singlet,
    p_dead = 1 - p_live - p_intermediate
  )
#write_tsv(gated_data, file = "../Input/20240124-fig-2-Dye-Order-gated-ht.tsv")
#print(gated_data)
```

MFI in the live and intermediate gates
```{r}
popMFI <- gs_pop_get_stats(gs, nodes = c("live", "intermediate"), type = pop.MFI) %>% 
  as_tibble() %>% 
  select(sample, pop, FSC = `FSC-H`, BL1 = `Fungalight Green-H`) %>% 
  pivot_wider(names_from = pop, values_from = c(FSC, BL1))
```

Combine the data and meta data
```{r}
final <- select(pData(dat), sample = name, everything()) %>% 
  left_join(select(gated_stats, -starts_with("n_")),  by = "sample") %>% 
  left_join(popMFI, by = "sample")

write_tsv(final, file = "../output/12.03-gated-output-20240214.tsv")
print(final %>% select(-sample))
```

# Compare with CFU
Read CFU data - these are averaged CFU provided by Hanxi. The raw counts data will be provided by her separately.
```{r}
cfu <- read_tsv("../input/20240214-17 Log H2O2 correlation Cg/20240219-log-h2o2-Cg-CFU-HT.tsv")
```

Calculate a simple 1-%live score for PI alone or SYTO9+PI, and merge with CFU
```{r}
PI <- final %>% 
  dplyr::filter(Dye == "PI") %>% 
  select(Date, Treatment, Method = Dye, Score = p_live)

FGL <- final %>% 
  dplyr::filter(Dye == "Both") %>% 
  mutate(
    wFGL = p_live * 2/log(BL1_live, 10)
  ) %>% 
  select(Date, Treatment, wFGL, FGL = p_live) %>% 
  pivot_longer(cols = c(wFGL, FGL), names_to = "Method", values_to = "Score")
  

all <- bind_rows(
  PI, FGL,
  mutate(cfu, Method = "CFU") %>% rename(Score = CFU)
) 
```
Plot
```{r}
all %>% 
  dplyr::filter(!Treatment %in% c("mock", "5C"), Method != "wFGL") %>% 
  mutate(Method = fct_relevel(Method, "CFU", "FGL", "PI")) %>% 
  ggplot(aes(x = Treatment, y = Score)) +
  geom_bar(stat = "summary", fun = "mean", fill = "gray") +
  #geom_point(size = 2, position = position_jitter(0.05)) + 
  stat_summary(fun.data = "mean_sdl", fun.args = list(mult = 1), 
               geom = "errorbar",  width = 0.1) +
  labs(x = bquote(H[2]*O[2]~(mM)), y = "% Live") +
  facet_wrap(~Method, nrow = 1) +
  theme_cowplot() +
  theme(strip.text = element_text(size = rel(1.2), face = 2),
        strip.background = element_blank())
ggsave("../output/20240323-for-R35-CFU-compared-with-FGL-and-PI.png", width = 5, height = 3)
```

```{r}
combined_data <- read_tsv("/space/htang5/Documents/FUN1_data_analysis/biol-4386-course-project-htang5/Methods Paper Writing/02-FungaLight-Distinguishing-Power/Input/20240219-PI-Both-Distinguishing-Power-gated-CFU-ht.tsv")
print(combined_data)
combined_data %>% 
  dplyr::filter(Dye %in% c("Both")) %>% 
  dplyr::filter(Treatment %in% c("0", "10", "100")) %>% 
  ggplot(aes(x = FGS, y = CFU_Survival)) +
  geom_point(aes(color = Treatment))

combined_data %>% 
  dplyr::filter(Dye %in% c("Both")) %>% 
  ggplot(aes(x = Treatment, y = FGS)) +
  geom_point(aes(color = Date))

combined_data %>% 
  dplyr::filter(Dye %in% c("Both")) %>% 
  ggplot(aes(x = Treatment, y = percent_inter)) +
  geom_point(aes(color = Date))
#ggsave("../output/20240104-PI-Dilutions-Gated_RP-Plot.png", width = 7, height = 5)
```


```{r}
combined_data$new_FGS <- 1*combined_data$percent_Live - 0.6*combined_data$percent_inter - combined_data$percent_Dead
pf1g <- combined_data %>%
  pivot_longer(cols = c(CFU_Survival, new_FGS),
               names_to = "Assay",
               values_to = "Survival") %>% 
  dplyr::filter(!(Dye %in% c("PI"))) %>% 
  dplyr::filter(!(Treatment %in% c("5C")))  %>%
     mutate(Assay = factor(Assay, levels = c("CFU_Survival", "new_FGS"), labels = c("CFU", "FGL"))) %>% 
  #mutate(Dilution = factor(Dilution, levels = c("250 x", "500 x", "1000 x", "2000 x"))) %>%
  ggplot(aes(x = Assay, y = Survival)) +
  facet_wrap(~ Treatment, nrow = 1, labeller = as_labeller(c("0" = "0 mM", "10" = "10 mM", "100" = "100 mM", "1000" = "1000 mM"))) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange", color = "red2",
               size = 0.8, position = position_nudge(x = 0.2)) +
  geom_point(stroke = 1, size = 2, position = position_jitter(0.01)) +
  scale_y_continuous() +
  theme_minimal()

pf1g +
  labs(x = "Assay Type", y = "Survival Quantification") +
  panel_border(color = "black", size = 1.5) +
  theme(axis.line = element_blank(),
        strip.background = element_blank(),
        axis.text.x = element_text(size = 20, face = "bold", color = "black"),  # Adjust size and face (bold)
        axis.text.y = element_text(size = 20, face = "bold", color = "black"),
        axis.title.x = element_text(size = 20, face = "bold"),
        axis.title.y = element_text(size = 20, face = "bold")
        )+ 
  theme(strip.text = element_text(size = 20, face = "bold")) 
ggsave("../output/20240301-Cg-Log-Gated_.png", width = 7, height = 5)
```




```{r}
combined_data$percent_PIDead <- 1 - (combined_data$n_PIDead / combined_data$n_singlet)
p <- combined_data %>%
  pivot_longer(cols = c(CFU_Survival, percent_PIDead),
               names_to = "PIAssay",
               values_to = "PISurvival") %>% 
  dplyr::filter(!(Dye %in% c("Both"))) %>% 
  dplyr::filter(!(Treatment %in% c("5C")))  %>%
     mutate(PIAssay = factor(PIAssay, levels = c("CFU_Survival", "percent_PIDead"), labels = c("CFU", "PI"))) %>% 
  #mutate(Dilution = factor(Dilution, levels = c("250 x", "500 x", "1000 x", "2000 x"))) %>%
  ggplot(aes(x = PIAssay, y = PISurvival)) +
  facet_wrap(~ Treatment, nrow = 1, labeller = as_labeller(c("0" = "0 mM", "10" = "10 mM", "100" = "100 mM", "1000" = "1000 mM"))) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange", color = "red2",
               size = 0.8, position = position_nudge(x = 0.2)) +
  geom_point(stroke = 1, size = 2, position = position_jitter(0.01)) +
  scale_y_continuous() +
  theme_minimal()

p +
  labs(x = "Assay Type", y = "Survival Quantification") +
  panel_border(color = "black", size = 1.5) +
  theme(axis.line = element_blank(),
        strip.background = element_blank(),
        axis.text.x = element_text(size = 20, face = "bold", color = "black"),  # Adjust size and face (bold)
        axis.text.y = element_text(size = 20, face = "bold", color = "black"),
        axis.title.x = element_text(size = 20, face = "bold"),
        axis.title.y = element_text(size = 20, face = "bold")
        )+ 
  theme(strip.text = element_text(size = 20, face = "bold")) 
ggsave("../output/20240301-Cg-Log-Gated-PI.png", width = 7, height = 5)
```

```{r}
pf1g <- combined_data %>%
  dplyr::filter(!(Dye %in% c("Both"))) %>% 
  dplyr::filter(!(Treatment %in% c("1000", "5C")))  %>%
  #mutate(Dilution = factor(Dilution, levels = c("250 x", "500 x", "1000 x", "2000 x"))) %>%
  ggplot(aes(x = Treatment, y = CFU_Survival)) +
  facet_wrap(~ Dye, scale = "free_y") +
  geom_boxplot() +
  geom_point(stroke = 1, size = 2, position = position_jitter(0.2)) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
pf1g +
  labs(x = "Hydrogen Peroxide Treatment Concentration", y = "CFU Survival") +
  panel_border(color = "black", size = 1.5) +
  theme(axis.line = element_blank(),
        strip.background = element_blank(),
        axis.text.x = element_text(size = 12, face = "bold", color = "black"),  # Adjust size and face (bold)
        axis.text.y = element_text(size = 16, face = "bold", color = "black"),
        axis.title.x = element_text(size = 16, face = "bold"),
        axis.title.y = element_text(size = 16, face = "bold")
        )+ 
  theme(strip.text = element_text(size = 16, face = "bold")) 
#ggsave("../output/20240104-PI-Dilutions-Gated_RP-Plot.png", width = 7, height = 5)
```




